---
title: "cc_2026_day14"
author: "Lyndsay Miles"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
30 Day Chart Challenge 2026 Day 14: Trade
https://rtais.wto.org/UI/PublicMaintainRTAHome.aspx
WTO Regional Trade Agreements Database, accessed 4/13/2026
```{r}
library(tidyverse)
library(lubridate)
library(ggthemes)
```

```{r}
df <- read.csv("data/AllRTAs.csv") 
```

```{r}
df_clean <- df %>%
  rename_with(~ tolower(gsub(" ", "", gsub(".", "", .x, fixed = TRUE), fixed = TRUE))) %>%
  mutate(
    date_parsed = lubridate::dmy(dateofentryintoforceg),
    # Fix two-digit years: if parsed year is after 2025, subtract 100
    date_parsed = if_else(year(date_parsed) > 2025, 
                          date_parsed - lubridate::years(100), 
                          date_parsed),
    year = lubridate::year(date_parsed)
  )

df_clean %>%
  count(year) %>%
  arrange(year)

# First recategorize status
df_clean <- df_clean %>%
  mutate(
    status_clean = case_when(
      status %in% c("In Force", "In force for at least one Party") ~ "In Force",
      status == "Inactive" ~ "Inactive",
      status %in% c("Early announcement-Under negotiation", 
                    "Early announcement-Signed ") ~ "In Negotiation",
      TRUE ~ "Other"
    ),
    year_inactive = lubridate::year(lubridate::dmy(inactivedate)),
    # Fix two digit years
    year_inactive = if_else(year_inactive > 2025,
                            year_inactive - 100,
                            year_inactive)
  )

# Check inactive year parsing
df_clean %>%
  filter(status_clean == "Inactive") %>%
  select(rtaname, year, year_inactive) %>%
  head(10)

```

```{r}
#how long does it take to get from notification to in force? 
df_clean %>%
  filter(status_clean == "In Force") %>%
  mutate(
    date_signature = lubridate::dmy(dateofsignatureg),
    # Fix two digit years
    date_signature = if_else(year(date_signature) > 2025,
                             date_signature - lubridate::years(100),
                             date_signature),
    days_to_force = as.numeric(date_parsed - date_signature),
    years_to_force = days_to_force / 365.25
  ) %>%
  filter(!is.na(years_to_force), years_to_force > 0) %>%  # remove negatives/NAs
  summarise(
    mean_years = round(mean(years_to_force), 1),
    median_years = round(median(years_to_force), 1),
    min_years = round(min(years_to_force), 1),
    max_years = round(max(years_to_force), 1),
    n = n()
  )

df_clean <- df_clean %>%
  mutate(composition_clean = case_when(
    rtacomposition %in% c("Bilateral", 
                          "Bilateral; All Parties are RTAs") ~ "Bilateral",
    rtacomposition == "Bilateral; One Party is an RTA" ~ "Bilateral (with Bloc)",
    rtacomposition %in% c("Plurilateral", 
                          "Plurilateral; One Party is an RTA") ~ "Plurilateral",
    TRUE ~ "Other"
  ))

df_clean %>% count(composition_clean)
```

```{r}
# Build year range
years <- tibble(year = 1975:2025)

# Count active agreements per year by composition
cumulative_by_comp <- years %>%
  mutate(
    bilateral = map_int(year, ~sum(
      df_clean$status_clean == "In Force" &
      df_clean$composition_clean == "Bilateral" &
      !is.na(df_clean$year) &
      df_clean$year <= .x
    )),
    bilateral_bloc = map_int(year, ~sum(
      df_clean$status_clean == "In Force" &
      df_clean$composition_clean == "Bilateral (with Bloc)" &
      !is.na(df_clean$year) &
      df_clean$year <= .x
    )),
    plurilateral = map_int(year, ~sum(
      df_clean$status_clean == "In Force" &
      df_clean$composition_clean == "Plurilateral" &
      !is.na(df_clean$year) &
      df_clean$year <= .x
    ))
  )

# Check tail
cumulative_by_comp %>% tail(5)

area_comp <- cumulative_by_comp %>%
  pivot_longer(cols = c(bilateral, bilateral_bloc, plurilateral),
               names_to = "composition",
               values_to = "count") %>%
  mutate(composition = recode(composition,
                              "bilateral" = "Bilateral",
                              "bilateral_bloc" = "Bilateral (with Bloc)",
                              "plurilateral" = "Plurilateral"),
         composition = factor(composition,
                              levels = c("Plurilateral",
                                         "Bilateral (with Bloc)",
                                         "Bilateral")))

# Calculate label positions at year 2008
label_positions <- area_comp %>%
  filter(year == 2008) %>%
  arrange(composition) %>%
  mutate(
    cumulative = cumsum(count),
    midpoint = cumulative - count / 2
  )
```

```{r}
#PLOT#
p <- ggplot(area_comp, aes(x = year, y = count, fill = composition)) +
  geom_area(alpha = 0.8, position = "stack") +
  scale_fill_manual(values = c(
    "Bilateral"             = "#1d3557",
    "Bilateral (with Bloc)" = "#457b9d",
    "Plurilateral"          = "#a8dadc"
  ),
                  breaks = c("Bilateral", "Bilateral (with Bloc)", "Plurilateral"))+
 
  # WTO annotation
  geom_vline(xintercept = 1994.5, linetype = "dashed",
             color = "grey50", linewidth = 0.4) +
  annotate("text", x = 1995, y = 350,
           label = "WTO founded (1995)",
           size = 4, hjust = 0, color = "grey40") +
  
  # Brexit annotation
  geom_vline(xintercept = 2021, linetype = "dashed",
             color = "grey50", linewidth = 0.4) +
  annotate("text", x = 2021, y = 300,
           label = "Brexit (2021)",
           size = 4, hjust = 1, color = "grey40") +

  scale_x_continuous(breaks = c(1970, 1980, 1990, 2000, 2010, 2020)) +
  scale_y_continuous(breaks = c(0, 100, 200, 300, 400)) +
  
  labs(
    fill=NULL,
    title = "50 Years of Regional Trade Agreements (RTA)",
    subtitle = "Cumulative RTA in force by composition | 1975-2025",
    x = NULL,
    y = NULL,
    caption = "Data: WTO Regional Trade Agreements Database\n#30DayChartChallenge Day 14 | Created by: Lyndsay Miles"
  ) +
  theme_economist() +
  theme(
    legend.position = "bottom",
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "grey85", linewidth = 0.3),
    plot.title = element_text(size = 22, face = "bold", hjust =0),
    plot.subtitle = element_text(size = 14, hjust = 0),
    plot.caption = element_text(hjust = 1, size = 12)
  )

p
ggsave("img/cc_2026_day14.png", p,
       width = 10, height = 6, 
       dpi = 300)
```
#Day14 of the #30DayChartChallenge | Topic: Trade 🤝. 
The number of regional trade agreements in force has grown from a handful to over 400 in the past 50 years. The composition has shifted too, with plurilateral agreements dominating early on, but now bilateral deals make up nearly 60% of all agreements in force. 

#Rstats #30DayChartChallenge #DataVisualization
Data: WTO Regional Trade Agreements Database (2026)
https://rtais.wto.org/UI/PublicMaintainRTAHome.aspx. 
