---
title: "cc_2026_day26"
author: "Lyndsay Miles"
date: "`r Sys.Date()`"
output: html_document
---
#Day26 #30DayChartChallenge Uncertainties: Trends
https://ghdx.healthdata.org/record/ihme-data/global-expected-health-spending-2023-2050-20251114


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(gganimate)
```

```{r}
df <- read_csv("data/IHME_EXPECTED_HEALTH_SPENDING_2023_2050_Y2025M11D06.CSV")

```

```{r}
names(df) #93 variables
glimpse(df)
unique(df$location_name) #216
unique(df$year) #2023-2050
unique(df$level) #Global, Country, World Bank Income Groups, GBD Super Regions
```

```{r}
df_asia <- df %>% 
  filter(location_name == "Southeast Asia, East Asia, and Oceania") %>%
  select(year, 
         ghes_per_cap_ppp_mean, 
         oop_per_cap_ppp_mean, 
         ppp_per_cap_ppp_mean, 
         
         )

  df_asia_long <- df_asia |>
    filter(year > 2024) |>
  pivot_longer(cols = c(ghes_per_cap_ppp_mean, oop_per_cap_ppp_mean, 
                         ppp_per_cap_ppp_mean),
               names_to = "spending_type",
               values_to = "value")
 
```


```{r}
#fonts
library(showtext)
font_add_google("Roboto", "roboto")
showtext_auto()

# clean labels for legend
asia_labels <- c(
  "ghes_per_cap_ppp_mean" = "Government",
  "oop_per_cap_ppp_mean"  = "Out-of-Pocket",
  "ppp_per_cap_ppp_mean"  = "Prepaid Private"
  
)

# set stack order — DAH at bottom, government on top
df_asia_long$spending_type <- factor(df_asia_long$spending_type,
  levels = c( 
             "oop_per_cap_ppp_mean",
             "ppp_per_cap_ppp_mean", "ghes_per_cap_ppp_mean")) 

p <- ggplot(df_asia_long, aes(x = year, y = value, 
                            fill = spending_type,
                            group = spending_type)) +
  geom_area(alpha = 0.85, color = NA, stat = "identity") +
  scale_fill_manual(values = c(
  "ghes_per_cap_ppp_mean" = "#2E9EC4",   
  "oop_per_cap_ppp_mean"  = "#E8B84B",   
  "ppp_per_cap_ppp_mean"  = "#A855C8"   
    
  ),
  labels = asia_labels) +
  scale_y_continuous(labels = scales::dollar) +
  scale_x_continuous(breaks = seq(2025, 2050, by = 5)) +
  labs(
    title = "Projected Health Spending Composition: SE Asia, E Asia & Oceania",
    subtitle = "In 2023 USD PPP | Year: {round(frame_along)}",
    x = "",
    y = "",
    fill = "",
    caption = "Source: IHME | #30DayChartChallenge | Created by: Lyndsay Miles"
  ) +
  transition_reveal(year) +
  theme_minimal(base_family = "roboto", base_size = 14)+
  theme(
    plot.background   = element_rect(fill = "#1C2B3A", colour = NA),
    panel.background  = element_rect(fill = "#1C2B3A", colour = NA),
    plot.title        = element_text(colour = "#FFFFFF", size = 18, margin = margin(b=15)),
    plot.subtitle     = element_text(colour = "#AAAAAA", size = 14),
    plot.caption      = element_text(colour = "#A0A0A0", size = 10, hjust = 1),
    axis.text         = element_text(colour = "#E8E8E8"),
    axis.title.y      = element_text(colour = "#E8E8E8", angle = 90, margin = margin(r = 10)),
    legend.text       = element_text(colour = "#E8E8E8", size = 10),
    legend.background = element_rect(fill = "#1C2B3A", colour = NA),
    legend.key        = element_rect(fill = "#1C2B3A", colour = NA),
    legend.position   = "right",
    legend.direction = "vertical",
    legend.spacing.x = unit(0.5, "cm"),
    legend.spacing.y = unit(0.3, "cm"),
    legend.key.width = unit(1, "cm"),
    panel.grid.major  = element_line(colour = "#2E4057", linewidth = 0.3),
    panel.grid.minor  = element_blank(),
    guides(fill = guide_legend(reverse = TRUE, nrow = 3)),
    plot.margin = margin(20, 150, 15, 25)  # top, right, bottom, left
  )
showtext_opts(dpi = 150)
animate(p, fps = 5, duration = 15, width = 1200, height = 700, renderer = gifski_renderer())
anim_save("img/cc_2026_day27.gif")
```

#Day27 of the #30DayChartChallenge | Animation (Uncertainties)
Continuing with the dataset I used for #Day26, my animation shows the projected healthcare spending in SE Asia, E Asia, and Oceania from 2025-2050. This was my first attempt at an animation outside of training, and #gganimate makes this an easier task. I picked this region because, outside of high-income countries, this region is projecting the greatest increase in government healthcare spending per capita (PPP). What surprised me was out-of-pocket spending is projected to be the largest and fastest-growing component. 

Data: IHME | #rstats #gganimate #dataviz #globalhealth 📈