---
title: "cc_2026_day26"
author: "Lyndsay Miles"
date: "`r Sys.Date()`"
output: html_document
---
#Day26 #30DayChartChallenge Uncertainties: Trends
https://ghdx.healthdata.org/record/ihme-data/global-expected-health-spending-2023-2050-20251114


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
```

```{r}
df <- read_csv("data/IHME_EXPECTED_HEALTH_SPENDING_2023_2050_Y2025M11D06.CSV")

```

```{r}
names(df) #93 variables
glimpse(df)
unique(df$location_name) #216
unique(df$year) #2023-2050
unique(df$level) #Global, Country, World Bank Income Groups, GBD Super Regions
```

```{r}
df_regions <- df %>% 
  #select the 7 GBD Super Regions
  filter(level == "GBD Super Regions") %>%
  #select Government Health Spending per person variables
  select(location_name, iso3, level, year, ghes_per_cap_ppp_mean, ghes_per_cap_ppp_lower, ghes_per_cap_ppp_upper)
```

```{r}
df_regions <- df_regions %>% filter(!location_name %in% "High-income", year > 2024)

ihme_colors <- c(
  "Central Europe, Eastern Europe, and Central Asia" = "#A05C7B",  # mauve/plum
  "Latin America and Caribbean"                      = "#C4A020",  # gold/ochre
  "North Africa and Middle East"                     = "#3D8B6E",  # teal green
  "South Asia"                                       = "#2E86AB",  # medium blue
  "Southeast Asia, East Asia, and Oceania"           = "#5B4FCF",  # purple-blue
  "Sub-Saharan Africa"                               = "#E05C2A"   # burnt orange
)
p<- ggplot(df_regions, aes(x = year, color = location_name, fill = location_name)) +
  geom_ribbon(aes(ymin = ghes_per_cap_ppp_lower, 
                  ymax = ghes_per_cap_ppp_upper), 
              alpha = 0.5, color = NA) +
  geom_line(aes(y = ghes_per_cap_ppp_mean)) +
  scale_y_continuous(labels = scales::dollar) +
  scale_x_continuous(
  limits = c(2025, 2050))+
  scale_color_manual(values = ihme_colors) +
  scale_fill_manual(values = ihme_colors)+
  labs(x = "",
       y = "2023 USD PPP",
       fill = "", color = "",
       title = "Current and Projected Government Healthcare Spending Per Capita",
       subtitle = "High income countries (projected: $7,601/person by 2050) excluded for scale",
       caption  = "Source: IHME | #30DayChartChallenge | Created by: Lyndsay Miles"  )+
  theme_minimal(base_size = 16)+
    theme(
      legend.text = element_text(color ="white"),
      legend.position = "bottom",
      axis.text   = element_text(colour = "white"),
      axis.title.y = element_text(colour = "white", angle = 90),
      plot.background   = element_rect(fill = "#1C2B3A", colour = NA),
      panel.background  = element_rect(fill = "#1C2B3A", colour = NA),
      plot.title = element_text(color = "white",
        size = 24, hjust = 0.5,margin = margin(b = 8)),
      plot.subtitle     = element_text(color = "white", size = 16,
                                     hjust = 0.5, margin = margin(b = 12)),
    plot.caption      = element_text(colour = "white", size = 10,
                                     hjust = 1, margin = margin(t = 10)),
    plot.margin       = margin(20, 25, 15, 25),
    panel.grid.major = element_line(colour = "#2E4057", linewidth = 0.3),
panel.grid.minor = element_blank()
  )

ggsave("img/cc_2026_day26.png", p, width = 12, height = 6.27, dpi = 300)
p
  
```
#Day26 of the #30DayChartChallenge | Trends (Uncertainties). When I thought about this prompt, I thought, what could be more uncertain than future spending? Here is a viz of projected government health care per capita spending (PPP) from 2025 to 2050, with confidence intervals, data courtesy of IHME's Global Expected Health Spending analysis. 📈