---
title: "cc_2026_day18"
author: "Lyndsay Miles"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
30 Day Chart Challenge 2026 Day 14: Trade
https://rtais.wto.org/UI/PublicMaintainRTAHome.aspx
WTO Regional Trade Agreements Database, accessed 4/13/2026
```{r}
library(tidyverse)
library(lubridate)
library(ggthemes)
```

```{r}
df <- read.csv("data/AllRTAs.csv") 
```

```{r}
df_clean <- df %>%
  rename_with(~ tolower(gsub(" ", "", gsub(".", "", .x, fixed = TRUE), fixed = TRUE))) %>%
  mutate(
    date_parsed = lubridate::dmy(dateofentryintoforceg),
    # Fix two-digit years: if parsed year is after 2025, subtract 100
    date_parsed = if_else(year(date_parsed) > 2025, 
                          date_parsed - lubridate::years(100), 
                          date_parsed),
    year = lubridate::year(date_parsed)
  )

df_clean %>%
  count(year) %>%
  arrange(year)

# First recategorize status
df_clean <- df_clean %>%
  mutate(
    status_clean = case_when(
      status %in% c("In Force", "In force for at least one Party") ~ "In Force",
      status == "Inactive" ~ "Inactive",
      status %in% c("Early announcement-Under negotiation", 
                    "Early announcement-Signed ") ~ "In Negotiation",
      TRUE ~ "Other"
    ),
    year_inactive = lubridate::year(lubridate::dmy(inactivedate)),
    # Fix two digit years
    year_inactive = if_else(year_inactive > 2025,
                            year_inactive - 100,
                            year_inactive)
  )

# Check inactive year parsing
df_clean %>%
  filter(status_clean == "Inactive") %>%
  select(rtaname, year, year_inactive) %>%
  head(10)

```

```{r}
#how long does it take to get from notification to in force? 
df_clean %>%
  filter(status_clean == "In Force") %>%
  mutate(
    date_signature = lubridate::dmy(dateofsignatureg),
    # Fix two digit years
    date_signature = if_else(year(date_signature) > 2025,
                             date_signature - lubridate::years(100),
                             date_signature),
    days_to_force = as.numeric(date_parsed - date_signature),
    years_to_force = days_to_force / 365.25
  ) %>%
  filter(!is.na(years_to_force), years_to_force > 0) %>%  # remove negatives/NAs
  summarise(
    mean_years = round(mean(years_to_force), 1),
    median_years = round(median(years_to_force), 1),
    min_years = round(min(years_to_force), 1),
    max_years = round(max(years_to_force), 1),
    n = n()
  )

df_clean <- df_clean %>%
  mutate(composition_clean = case_when(
    rtacomposition %in% c("Bilateral", 
                          "Bilateral; All Parties are RTAs") ~ "Bilateral",
    rtacomposition == "Bilateral; One Party is an RTA" ~ "Bilateral (with Bloc)",
    rtacomposition %in% c("Plurilateral", 
                          "Plurilateral; One Party is an RTA") ~ "Plurilateral",
    TRUE ~ "Other"
  ))

df_clean %>% count(composition_clean)
```

```{r}
library(tidyverse)
# Step 1: Filter to Asia-Pacific agreements
df_asia <- df_clean %>%
  mutate(year = lubridate::year(lubridate::dmy(dateofentryintoforceg)),
         year = if_else(year > 2025, year - 100, year)) %>%
  mutate(composition_clean = case_when(
    rtacomposition %in% c("Bilateral", 
                          "Bilateral; All Parties are RTAs") ~ "Bilateral",
    rtacomposition == "Bilateral; One Party is an RTA" ~ "Bilateral (with Bloc)",
    rtacomposition %in% c("Plurilateral", 
                          "Plurilateral; One Party is an RTA") ~ "Plurilateral",
    TRUE ~ "Other"
  )) %>%
  filter(str_detect(region, regex("east asia|oceania", ignore_case = TRUE)),
         status_clean == "In Force") %>%
  mutate(year = lubridate::year(lubridate::dmy(dateofentryintoforceg)),
         year = if_else(year > 2025, year - 100, year)) %>%
  drop_na(year)

# Step 2: Parse region into pairs
region_pairs <- df_asia %>%
  select(rtaname, region, composition_clean) %>%
  mutate(region = str_split(region, ";\\s*")) %>%
  unnest(region) %>%
  mutate(region = str_trim(region))

# Step 3: Create from-to pairs
chord_data <- df_asia %>%
  select(rtaname, region) %>%
  mutate(
    region_list = map(str_split(region, ";\\s*"), str_trim),
    from = map_chr(region_list, ~.x[1]),
    to = map_chr(region_list, ~if(length(.x) >= 2) .x[2] else .x[1])
  ) %>%
  count(from, to) %>%
  arrange(desc(n))

# Step 4: Simplify small regions into "Other"
chord_data_clean <- chord_data %>%
  mutate(
    from = case_when(
      from %in% c("Africa", "Central Asia", "Caribbean",
                  "Middle East", "Central America",
                  "North America") ~ "Other",
      TRUE ~ from
    ),
    to = case_when(
      to %in% c("Africa", "Central Asia", "Caribbean",
                "Middle East", "Central America",
                "North America") ~ "Other",
      TRUE ~ to
    )
  ) %>%
  group_by(from, to) %>%
  summarise(n = sum(n), .groups = "drop") %>%
  arrange(desc(n))

chord_data_clean

chord_data_alluvial <- chord_data_clean %>%
  filter(from != "Other", to != "Other") %>%
  filter(!(from == to)) %>%  # remove self-loops
  mutate(
    from_label = paste0(from, " "),   # add trailing space to from
    to_label = paste0(" ", to)        # add leading space to to
  )
```

```{r}
# Step 5: Build chord diagram
library(ggalluvial)
library(ggtext)


windowsFonts(georgia = windowsFont("Georgia"))

ggplot(chord_data_alluvial,
       aes(axis1 = from_label, axis2 = to_label, y = n)) +
  geom_alluvium(aes(fill = from), 
                alpha = 0.8, 
                width = 1/3) +
  geom_stratum(width = 1/3, fill = "white", color = NA) +
  geom_text(stat = "stratum",
            aes(label = after_stat(stratum)),
            size = 6, family = "georgia") +
  scale_fill_manual(values = c(
    "East Asia"     = "#4a90a4",
    "Oceania"       = "#c9a84c",
    "Europe"        = "#e8a598",
    "South America" = "#7ab648",
    "West Asia"     = "#9b6ea8"
  )) +
  labs(
    title = "<span style='color:#4a90a4'>Asia-Pacific's Web of Trade Relationships</span>",
    subtitle = "Regional trade agreements in force involving East Asia or Oceania | WTO 2026",
    x = NULL, y = NULL,
    caption = "Data: WTO Regional Trade Agreements Database | Intra-regional agreements removed for clarity.\n#30DayChartChallenge Day 18 | Created by: Lyndsay Miles"
  ) +
  theme_void(base_size = 14) +
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    panel.border = element_blank(),
    panel.background = element_rect(fill = "white", color = NA),
    plot.title = element_markdown(family = "georgia", size = 24,
                              face = "bold", hjust = 0.5),
    plot.subtitle = element_text(family = "georgia", size = 14,
                                 color = "grey40", hjust = 0.5),
    plot.caption = element_text(family = "georgia", size = 10,
                                hjust = 1, color = "grey50"),
    legend.position = "none",
    plot.margin = margin(30, 20, 20, 20)
  )

ggsave("img/cc_2026_day18.png", 
       width = 10, height = 8, 
       dpi = 300,
       bg = "white")
```

#Day18 of the #30DayChartChallenge | Theme Day: South China Morning Post 🌏

Today I went back to the WTO Regional Trade Agreements data from #Day14, filtered to agreements involving East Asia or Oceania, and visualized with SCMP's style.

The Sankey diagram shows which regions are connected through formal trade relationships. East Asia dominates — its largest connections are with Europe and Oceania. South America also has a strong trading relationship with East Asia (10 agreements). There were 4 agreements involving North America, but I left these, as well as intra-regional agreements, out for clarity.

#Rstats #30DayChartChallenge #DataVisualization #TradePolicy
Data: WTO Regional Trade Agreements Database (2026)
https://rtais.wto.org/UI/PublicMaintainRTAHome.aspx. 
