---
title: "cc_2026_day28"
author: "Lyndsay Miles"
date: "`r Sys.Date()`"
output: html_document
---
#Day28 #30DayChartChallenge Modeling (Uncertainties)
https://github.com/rfordatascience/tidytuesday/blob/main/data/2026/2026-03-10/readme.md
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(ggplot2)
library(lme4)
```

```{r}
absolute_judgements <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2026/2026-03-10/absolute_judgements.csv')
pairwise_comparisons <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2026/2026-03-10/pairwise_comparisons.csv')
respondent_metadata <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2026/2026-03-10/respondent_metadata.csv')


```

```{r}
# how many unique respondents?
n_distinct(absolute_judgements$response_id)

# how much missing demographic data?
respondent_metadata |> 
  summarise(across(everything(), ~sum(is.na(.))))

# distribution of estimates per term
absolute_judgements |>
  group_by(term) |>
  summarise(mean = mean(probability),
            sd = sd(probability)) |>
  arrange(desc(sd))

```

```{r}
df <- absolute_judgements |>
  left_join(respondent_metadata, by = "response_id")
```

```{r}
#check for nas
df_complete <- df |>
  filter(!is.na(age_band), 
         !is.na(english_background))

df_complete <- df_complete |>
  mutate(english_binary = case_when(
    english_background == "English is my first language" ~ "First language",
    TRUE ~ "Not first language"
  ))

# check how many respondents remain
n_distinct(df$response_id)

#size sample imbalance?
df_complete |>
  distinct(response_id, english_binary) |>
  count(english_binary)

df_complete |>
  distinct(response_id, english_binary) |>
  count(english_binary)
```


```{r}
model <- lmer(probability ~ term * english_binary + 
                  
                 (1|response_id), 
               data = df_complete,
               na.action = na.exclude)

summary(model)
```

```{r}
# extract fixed effects
coefs <- as.data.frame(summary(model)$coefficients)
coefs$term <- rownames(coefs)

interactions <- coefs |>
  filter(grepl(":", term)) |>
  rename(estimate = Estimate,
         se = `Std. Error`,
         t_value = `t value`) |>
  mutate(
    prob_term = gsub("term(.+):english_binary.*", "\\1", term),
    eng_group = case_when(
      grepl("Not first language", term) ~ "Not first language"
    ),
    significant = abs(t_value) > 2
  ) |>
  select(prob_term, eng_group, estimate, se, t_value, significant)
head(interactions)
```

```{r}
# order terms by mean estimate for cleaner layout
term_order <- interactions |>
  arrange(estimate) |>
  pull(prob_term)

interactions <- interactions |>
  mutate(prob_term = factor(prob_term, levels = term_order))
```

```{r}
p <- ggplot(interactions, aes(x = estimate, y = prob_term)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey60") +
  geom_errorbar(aes(xmin = estimate - 1.96 * se,
                     xmax = estimate + 1.96 * se),
                 height = 0.2, alpha = 0.5, color = "#9B59B6") +
  geom_point(aes(shape = significant, size = significant),
             color = "#9B59B6") +
  scale_shape_manual(values = c("TRUE" = 19, "FALSE" = 1)) +
  scale_size_manual(values = c("TRUE" = 3, "FALSE" = 2)) +
  coord_cartesian(xlim = c(-6, 6)) +
  labs(
    title = "Are probability phrase interpretations consistent across English language backgrounds?",
    subtitle = "Non-native speakers assign systematically different probabilities to select phrases",
    x = "Difference in probability estimate relative to native English speakers (percentage points)",
    y = NULL,
    caption = "Filled points indicate |t| > 2\nMixed-effects model, n=5,174 | Data: #tidytuesday | Kucharski probability phrases survey | #30DayChartChallenge Day 28"
  ) +
  ggthemes::theme_fivethirtyeight(base_size=16)+
  theme(
    legend.position = "none",
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 22, face = "bold"),
    plot.subtitle = element_text(),
    plot.title.position = "plot"
  )

ggsave(filename = "img/cc_2026_day28.png",
       plot = p,
       width = 12,
       height = 7,
       dpi = 300)
```
It's #Day28 of the #30DayChartChallenge and the prompt is Modeling (Uncertainties). The data comes from Adam Kucharski's online quiz to better understand how people interpret common probability phrases, courtesy of #tidytuesday. 

I was curious if probability phrase interpretations would be consistent across English language backgrounds, and used a linear mixed-effects model to test whether non-native English speakers interpret common probability phrases differently than native speakers. For certain phrases, this appears to be true. For example, non-native speakers (n=679) tend to assign higher probabilities to "chance" language like "Remote Chance" and "Chances are Slight," compared to native speakers (N = 3,801).
However, there are some caveats - this is an initial analysis and given more time I would want to explore some aspects further. Want to take the quiz yourself? It's at https://probability.kucharski.io/.
#30DayChartChallenge #dataviz #rstats #publichealth #lme4



