---
title: "cc_2026_day25"
author: "Lyndsay Miles"
date: "`r Sys.Date()`"
output: html_document
---
#Day25 #30DayChartChallenge: Space (Uncertainties)


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(lubridate)
library(showtext)
library(ggtext)
library(rtrek)
library(trekcolors)
library(trekfont)
library(ggrepel)
```

```{r}
#https://datasets.imdbws.com/
ratings <- read_tsv("data/title.ratings.tsv.gz",  na = "\\N")
episodes <- read_tsv("data/title.episode.tsv.gz", na = "\\N")
```

```{r}
#DS9 IMDB data
# Filter to DS9 episodes only and join ratings
ds9 <- episodes |>
  filter(parentTconst == "tt0106145") |>
  left_join(ratings, by = "tconst") |>
  mutate(
    season  = as.integer(seasonNumber),
    episode = as.integer(episodeNumber)
  ) |>
  filter(!is.na(season), !is.na(episode), !is.na(averageRating)) |>
  select(
    tconst,
    season,
    episode,
    rating = averageRating,
    votes  = numVotes
  ) |>
  arrange(season, episode)
View(ds9)
```

```{r}
ds9_seasons <- ds9 |>
  group_by(season) |>
  summarise(
    avg_rating  = mean(rating),
    sd_rating   = sd(rating),
    n_episodes  = n(),
    total_votes = sum(votes)
  ) |>
  #use t-distribution to calculate 95% confidence intervals because # of episodes per season is less than 30. Originally wanted to visualize the CI as a halo around each circle, but the ranges were too small to be visually interesting
  mutate(
    se      = sd_rating / sqrt(n_episodes),
    margin  = qt(0.975, df = n_episodes - 1) * se,
    ci_low  = avg_rating - margin,
    ci_high = avg_rating + margin
  )

#check
ds9_seasons |>
  select(season, n_episodes, avg_rating, ci_low, ci_high, margin)
```

```{r}
#palette
lcars_cols <- trekcolors::trek_pal("lcars_2369")[1:7]

names(lcars_cols) <- as.character(1:7)
```

```{r}
font_add("ds9title",
  system.file("fonts/FederationDS9Title.ttf", package = "trekfont"))
font_add("fedbold",
  system.file("fonts/FederationBold.ttf", package = "trekfont"))

library(showtext)
showtext_auto()
showtext_opts(dpi = 300)
p<-
ggplot(ds9_seasons, aes(x = factor(season), y = avg_rating)) +

# core bubble: fixed size
  geom_point(
    aes(colour = factor(season)),
    size = 14,
    alpha = 0.90
  ) +

  # season labels
  geom_text(
  aes(label = paste0("S", season)),
  family = "fedbold",
  colour = "white",
  size = 4,
  fontface = "bold"
)+
  
  # S1 text box in lower left
annotate(
  "text",
  x = 0.6, y = 6.45,
  label = "A rocky start: DS9's premiere\nachieved an 18.8 Nielsen rating —\nthe highest in Trek history. But\nserialized storytelling, a stationary\nsetting, and competition from TNG\nkept early ratings middling.",
  colour = "#9999CC",
  size = 4,
  hjust = 0,
  lineheight = 1.4
) +

# leader line from text box to S1 bubble
annotate(
  "segment",
  x = 1, xend = 1,
  y = 6.65, yend = 6.88,
  colour = "#9999CC",
  linewidth = 0.3,
  linetype = "dashed",
  alpha = 0.6
) +

# S4 annotation above bubble
annotate(
  "text",
  x = 4, y = 7.95,
  label = "First Emmy nomination",
  colour = "#CC6699",
  size = 4,
  hjust = 0.5
) +

# S7 annotation above bubble
annotate(
  "text",
  x = 6.8, y = 7.95,
  label = "4 Primetime Emmy Awards\nearned across the series",
  colour = "#FF9900",
  size = 4,
  hjust = 0.5,
  lineheight = 1.3
) +

  scale_colour_manual(values = lcars_cols) +
  guides(colour = "none", size = "none") +

  scale_y_continuous(
  limits = c(6.2, 8.1),  
  breaks = seq(6.5, 8.0, 0.5)
)+

  theme_void(base_size = 16) +
  theme(
    plot.background   = element_rect(fill = "black", colour = NA),
    panel.background  = element_rect(fill = "black", colour = NA),
    axis.text.x  = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.y       = element_text(colour = "grey60", size = 16),
    panel.grid.major  = element_line(colour = "grey15", linewidth = 0.3),
    plot.title = element_text(family = "ds9title", colour = "white",
                           size = 32, hjust = 0.5,
                           margin = margin(b = 4)),
    plot.subtitle     = element_text(colour = "grey50", size = 20,
                                     hjust = 0.5, margin = margin(b = 12)),
    plot.caption      = element_text(colour = "grey40", size = 14,
                                     hjust = 1, margin = margin(t = 10)),
    plot.margin       = margin(20, 25, 15, 25)
  ) +

  
  labs(
    title    = "DEEP SPACE NINE",
    subtitle = "Average episode IMDB rating by season",
    caption  = "Source: IMDb Non-Commercial Datasets | #30DayChartChallenge | Created by: Lyndsay Miles"
  )

ggsave("img/cc_2026_day25.png", p, width = 12, height = 6.27, dpi = 300)
p
```
#Day25 of the #30DayChartChallenge: Space (Uncertainties). 🌌 Did you know there's a Star Trek palette package in R? And a Star Trek font package? Now you know — you're welcome. There's also the rtrek package which I didn't use here, but would be great for text analysis of episode dialogue. Today's chart traces Deep Space Nine's journey across 7 seasons of average IMDb episode ratings.
Data Source: IMDB. R packages: rtrek, trekcolors, trekfont, tidyverse, ggplot2
#30DayChartChallenge #dataviz
