---
title: "cc_2026_day3"
author: "Lyndsay Miles"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
30 Day Chart Challenge 2026 Day 4: Slope

```{r}
library(tidyverse)
library(ggplot2)
library(RColorBrewer)
library(extrafont)
library(showtext)
library(WDI)
```

```{r}
spending_purpose <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2026/2026-04-21/spending_purpose.csv')
financing_schemes <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2026/2026-04-21/financing_schemes.csv')
health_spending <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2026/2026-04-21/health_spending.csv')
```
Cleaning 
```{r}
pop_2022 <- WDI(
  indicator = "SP.POP.TOTL",
  country   = c("US", "CA", "MX"),  
  start     = 2022,
  end       = 2022
)

pop_2018 <- WDI(
  indicator = "SP.POP.TOTL",
  country   = c("US", "CA", "MX"),  
  start     = 2018,
  end       = 2018
)

pop <- bind_rows(pop_2022, pop_2018) %>% 
  rename(iso3_code = iso3c, tot_n = SP.POP.TOTL)%>% 
  select(iso3_code, year, tot_n)
```


```{r}
df <- spending_purpose %>% 
  filter(spending_purpose == "Preventive care",
         unit == "constant 2023 US$",
         iso3_code %in% c("USA", "MEX", "CAN"),
         year %in% c("2018", "2022"))%>% 
  left_join(pop, by = c("iso3_code", "year"))%>%
  mutate(percap_spend = value/tot_n)
```

Add dyslexic-friendly fonts
```{r}
font_add_google("Atkinson Hyperlegible", "atkinson")
showtext_auto()
```

```{r}
plot <- ggplot(
  data = df,
  aes(y = percap_spend, x = year, color = iso3_code, group = iso3_code))+
  geom_point()+
  geom_line(linewidth=1)+
  scale_x_continuous(
  breaks = c(2018, 2022),
  labels = c("2018", "2022")
)+
  theme_void()+
  scale_color_manual(values = c("#019875FF", "#FF847CFF", "#96281BFF"),
                     labels = c("CAN" = "Canada", "MEX" = "Mexico", "USA" = "United States"))+
  labs(
    title    = "North America Is Investing More in Prevention, But Not Equally",
    subtitle = "Change in Per Capita Spending on Preventive Care Between 2018 and 2022",
    caption  = "Source: WHO Global Health Expenditure Database (GHED) | #30DayChartChallenge Day 4 (slope)| #tidytuesday\nCreated by: Lyndsay Miles",
    color    = NULL
  ) +
  theme(
    text                  = element_text(family = "atkinson"),
    plot.title            = element_text(face = "bold", size = 16, hjust = 0,
                                         margin = margin(b = 6)),
    plot.subtitle         = element_text(size = 10, hjust = 0,
                                         margin = margin(b = 16)),
    plot.caption          = element_text(size = 7, color = "gray50", hjust = 0,
                                         margin = margin(t = 12)),
    plot.title.position   = "plot",
    plot.caption.position = "plot",
    plot.background       = element_rect(fill = "#f5ebe0", color = NA),
    axis.text.x = element_text(size = 10, face = "bold"),
    axis.text.y           = element_text(size = 10, face = "bold", hjust = 1,
                                         margin = margin(r = 8)),
    legend.position       = "bottom",
    legend.text           = element_text(size = 9),
    plot.margin           = margin(20, 20, 20, 20)
  ) +
  guides(color = guide_legend(override.aes = list(size = 3)))

plot
```
Day 4 of the #30DayChartChallenge: Slope — visualizing change over time.
Sticking with my theme, I returned to the WHO GHED spending purpose data from Day 1 to examine per capita preventive care spending changes across North America between 2018 and 2022.

There's a notable increase, especially for the USA but I suspect this reflects COVID-19 prevention spending rather than a structural shift.

What slope charts don't show well is relative change. Canada and Mexico each saw ~30% increases in per capita spending, but because the USA started so much higher, their ~70% increase crowds out the others.
