---
title: "cc_2026_day15"
author: "Lyndsay Miles"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
30 Day Chart Challenge 2026 Day 15: Correlation

```{r}
library(tidyverse)
library(ggthemes)
library(mosaicData)
library(ggplot2)
library(patchwork)
```

```{r}
df <- mosaicData::Whickham
```

```{r}
df %>% 
  group_by(smoker) %>%
  summarise(
    n = n(),
    pct_alive = round(mean(outcome == "Alive") * 100, 1)
  ) 
#Correlation - when age is not taken into account, it appears that smoking is protective; 76.1% of smokers are alive compared t0 68.6 non smokers

```

```{r}
df %>%
  summarise(
    min_age = min(age),
    max_age = max(age),
    mean_age = mean(age),
    median_age = median(age)
  )

# And a histogram to see the shape
ggplot(df, aes(x = age)) +
  geom_histogram(binwidth = 5, fill = "#457b9d", color = "white") +
  theme_minimal()
```

Causation
First bar chart includes dots at random, indicating individuals colored by age group but grouped by smokers and non-smokers. Second one redistributes those folks into their age groups, showing the truth about relationship between mortality and smoking.
```{r}
windowsFonts(georgia = windowsFont("Georgia"))
# Individual level data with all variables
dot_data <- df %>%
  mutate(
    age_group = cut(age,
                    breaks = c(18, 24, 44, 54, 64, 84),
                    labels = c("18-24", "25-44", "45-54", "55-64", "65+"),
                    include.lowest = TRUE),
    dead = if_else(outcome == "Dead", 1, 0),
    smoker_label = if_else(smoker == "Yes", "Smoker", "Non-Smoker")
  )

bar_data <- dot_data %>%
  group_by(smoker_label) %>%
  summarise(
    pct_dead = round(mean(dead) * 100, 1),
    .groups = "drop"
  )
dot_data_chart1 <- dot_data %>%
  left_join(bar_data, by = "smoker_label") %>%
  mutate(
    y_pos = runif(n(), 0, pct_dead),
    x_jitter = runif(n(), -0.25, 0.25),  # change -0.25, 0.25 to -0.15, 0.15
    x_pos = if_else(smoker_label == "Smoker", 2 + x_jitter, 1 + x_jitter)
  )
#Change -0.25, 0.25 to -0.15, 0.15 on the x_jitter line!

# Summary for bar heights
# Make bar_data numeric
bar_data_numeric <- bar_data %>%
  mutate(x_numeric = if_else(smoker_label == "Smoker", 2, 1))

p1 <- ggplot() +
  # Bars using numeric x
  geom_col(data = bar_data_numeric,
           aes(x = x_numeric, y = pct_dead),
           fill = "#a8dadc", alpha = 0.4, width = 0.5) +
  # Dots using numeric x_pos
  geom_point(data = dot_data_chart1,
           aes(x = x_pos, y = y_pos, color = age_group),
           alpha = 0.5, size = 1.5) +
scale_color_manual(values = c(
  "18-24" = "#2a9d8f",
  "25-44" = "#57cc99",
  "45-54" = "#e9c46a",
  "55-64" = "#f4a261",
  "65+"   = "#e63946"
)) +
theme(legend.position = "bottom",
      legend.title = element_blank())+
  # Labels using numeric x
  geom_text(data = bar_data_numeric,
            aes(x = x_numeric, y = pct_dead + 1.5,
                label = paste0(pct_dead, "%")),
            size = 4, fontface = "bold", color = "#1d3557") +
  scale_x_continuous(breaks = c(1, 2),
                   labels = c("Non-Smoker\n(n=732)", "Smoker\n(n=582)"),
                     limits = c(0.5, 2.5)) +
  labs(
    title = "The Simpson's Paradox: Smokers Outlive Non-Smokers?",
    subtitle = "20-year all-cause mortality proportion | Dot count reflects sample size",
    color = "Age Group",
    x = NULL, y = NULL
    
  ) +
  theme_minimal() +
    theme(
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    panel.grid = element_blank(),
    axis.text.y = element_blank(),
    axis.text.x = element_text(size = 11, face = "bold", family = "georgia"),
    plot.title = element_text(size = 16, face = "bold", family = "georgia"),
    plot.subtitle = element_text(size = 11, family = "georgia"),
    plot.caption = element_text(hjust = 1, size = 8, family = "georgia"),
    legend.text = element_text(family = "georgia"),
    legend.position = "bottom"
      )

p1

#Day 16: Correlation. % who died at 20-year follow-up by age group
# Summary for bar heights by age group and smoker status
bar_data2 <- dot_data %>%
  group_by(age_group, smoker_label) %>%
  summarise(pct_dead = round(mean(dead) * 100, 1), .groups = "drop")

age_levels <- c("18-24", "25-44", "45-54", "55-64", "65+")

set.seed(100)

dot_data_chart2 <- dot_data %>%
  left_join(bar_data2, by = c("age_group", "smoker_label")) %>%
  mutate(
    age_numeric = as.numeric(factor(age_group, levels = age_levels)),
    x_offset = if_else(smoker_label == "Smoker", 0.22, -0.22),
    x_jitter = runif(n(), -0.15, 0.15),
    x_pos = age_numeric + x_offset + (x_jitter * 0.8),
    y_pos = runif(n(), 0, pct_dead * 0.9)
    #y_pos = runif(n(), 0, pct_dead)
  )

bar_data2_numeric <- bar_data2 %>%
  mutate(
    age_numeric = as.numeric(factor(age_group, levels = age_levels)),
    x_offset = if_else(smoker_label == "Smoker", 0.22, -0.22),
    x_pos = age_numeric + x_offset
  )

p2 <- ggplot() +
  geom_col(data = bar_data2_numeric,
           aes(x = x_pos, y = pct_dead, fill = smoker_label),
           alpha = 0.3, width = 0.4) +
  scale_fill_manual(values = c(
    "Non-Smoker" = "#2a9d8f",
    "Smoker"     = "#e63946"
  )) +
  geom_point(data = dot_data_chart2,
             aes(x = x_pos, y = y_pos, color = smoker_label),
             alpha = 0.5, size = 1.5) +
  geom_text(data = bar_data2_numeric,
          aes(x = x_pos, y = pct_dead + 2,
              label = paste0(pct_dead, "%")),
          size = 3, fontface = "bold", color = "#1d3557") +
  scale_color_manual(values = c(
    "Non-Smoker" = "#2a9d8f",
    "Smoker"     = "#e63946"
  )) +
  scale_x_continuous(breaks = 1:5,
                     labels = age_levels) +
  labs(
    title = "Stratifying By Age Reveals the Truth.",
    subtitle = "20-year all-cause mortality proportion, by age group",
    x = NULL, y = NULL,
    color = NULL,
    fill = NULL,
    caption = "Data: Whickham Study, 1972-1974 (Appleton et al. 1996) | #30DayChartChallenge Day 15 & 16 (Correlation & Causation)\nCreated by: Lyndsay Miles"
  ) +
  theme_minimal() +
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    panel.grid = element_blank(),
    axis.text.y = element_blank(),
    axis.text.x = element_text(size = 11, face = "bold", family = "georgia"),
    plot.title = element_text(size = 22, face = "bold", family = "georgia"),
    plot.subtitle = element_text(size = 14, family = "georgia"),
    plot.caption = element_text(hjust = 1, size = 10, family = "georgia"),
    legend.text = element_text(family = "georgia"),
    legend.position = "bottom"
      )

p2

p3 <- p1/p2
ggsave("img/cc_2026_day15.png", p1,
       width = 10, height = 6, 
       dpi = 300)

ggsave("img/cc_2026_day16.png", p2,
       width = 10, height = 6, 
       dpi = 300)
```

```{r}
ggsave("img/cc_2026_day15_16.png", p3,
       width = 10, height = 6, 
       dpi = 300)
p3
```

Initial analysis of the data on women showed that over the 20-year follow-up, 31.4% of non-smokers had died, while only 23.9% of smokers had died