---
title: "cc_2026_day3"
author: "Lyndsay Miles"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
30 Day Chart Challenge 2026 Day 5: Experimental

```{r}
library(tidyverse)
library(ggplot2)
library(RColorBrewer)
library(extrafont)
library(showtext)
library(ggbeeswarm)
```

```{r}
spending_purpose <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2026/2026-04-21/spending_purpose.csv')
financing_schemes <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2026/2026-04-21/financing_schemes.csv')
health_spending <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2026/2026-04-21/health_spending.csv')
```
Exploration
```{r}
spending_purpose %>% filter(spending_purpose == "Preventive care", unit == "% of current health expenditure")%>% 
  group_by(year, iso3_code) %>% 
  summarize(N = n())
#142 unique counties
#801 rows
```

Cleaning
```{r}
library(countrycode)

df <- spending_purpose %>%
  filter(spending_purpose == "Preventive care",
         unit == "% of current health expenditure",
         year %in% c(2016:2023)) %>%
  mutate(region = countrycode(iso3_code, 
                              origin      = "iso3c", 
                              destination = "region"),
         region = if_else(iso3_code == "MEX", "North America", region))

# How many countries per region?
df %>% 
  distinct(iso3_code, region) %>% 
  count(region)

# Any NAs?
df %>% filter(is.na(region))
```

```{r}
font_add_google("Atkinson Hyperlegible", "atkinson")
showtext_auto()

# Highlight USA, CAN, MEX
df <- df %>%
  mutate(
    highlight = case_when(
      iso3_code == "USA" ~ "United States",
      iso3_code == "CAN" ~ "Canada",
      iso3_code == "MEX" ~ "Mexico",
      TRUE               ~ "Other"
    ),
    highlight = factor(highlight, 
                       levels = c("United States", "Canada", 
                                  "Mexico", "Other"))
  )
```

```{r}
ggplot(df, aes(x = region, y = value)) +
  geom_beeswarm(
    data  = df %>% filter(highlight == "Other"),
    size  = 1.2, alpha = 0.5, cex = 1.5, 
    color = "#c0c0c0"
  ) +
  geom_beeswarm(
    data  = df %>% filter(highlight != "Other"),
    aes(color = highlight),
    size  = 2, alpha = 0.9, cex = 1.5
  ) +
  scale_color_manual(
    values = c(
      "United States" = "#a6611a",
      "Canada"        = "#018571",
      "Mexico"        = "#80cdc1"
    )
  )+
  scale_x_discrete(labels = function(x) stringr::str_wrap(x, width = 12)) +
  labs(
    title   = "How Does Preventive Care Spending in North America Compare With Other Regions?",
    subtitle = "% of current health expenditure spent on preventive care, 2016–2023",
    caption  = "Source: WHO Global Health Expenditure Database (GHED)\n#30DayChartChallenge Day 5 (Experimental) | #tidytuesday\nCreated by: Lyndsay Miles",
    x        = NULL,
    y        = NULL,
    color    = NULL
  ) +
  theme_minimal() +
  theme(
    text                  = element_text(family = "atkinson"),
    plot.title            = element_text(face = "bold", size = 14, hjust = 0,
                                         margin = margin(b = 6)),
    plot.subtitle         = element_text(size = 10, hjust = 0,
                                         margin = margin(b = 16)),
    plot.caption          = element_text(size = 7, color = "gray50", hjust = 0,
                                         margin = margin(t = 12)),
    plot.title.position   = "plot",
    plot.caption.position = "plot",
    plot.background       = element_rect(fill = "#f5ebe0", color = NA),
    panel.background      = element_rect(fill = "#f5ebe0", color = NA),
    panel.grid.major.x    = element_blank(),
    panel.grid.major.y    = element_line(color = "#e0d5c8"),
    axis.text.x           = element_text(size = 8, face = "bold"),
    axis.text.y           = element_text(size = 9),
    legend.position       = "bottom",
    legend.text           = element_text(size = 9),
    plot.margin           = margin(20, 20, 5, 20)
  ) +
  guides(color = guide_legend(override.aes = list(size = 4)))
```
Day 5 of the #30DayChartChallenge (Experimental) 🐝

For today's prompt I tried out a visualization new to me, the beeswarm plot. It is essentially a specialized scatter plot that shows individual data points without overlap, revealing both density and distribution.

I used the #Rstats ggbeeswarm package to look at preventive care spending as a % of current health expenditure across world regions from 2016–2023, highlighting the three North American countries I followed in earlier charts.

North American spending on preventive care clusters tightly at the bottom while other regions show much wider variation in how much they prioritize prevention. Sub-Saharan Africa in particular shows striking spread with some countries dedicating well over 30% of health expenditure to prevention while others spend very little.


