---
title: "cc_2026_day13"
author: "Lyndsay Miles"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
30 Day Chart Challenge 2026 Day 13: Ecosystems
CalEnviroScreen4.0 Data
Definitions:
Cardiovascular disease: Age-adjusted rate of emergency department visits for heart attacks per 10,000 
CES 4.0 Score    : CalEnviroScreen Score, Pollution Score multiplied by Population Characteristics Score 
Total Population: 2019 ACS population estimates in census tracts
```{r}
library(tidyverse)
library(readxl)
```

```{r}
df <- read_excel("C:/Users/lymil/OneDrive/Desktop/R_Projects/calenviro_example/data/calenviroscreen40resultsdatadictionary_F_2021.xlsx", sheet = 1, na = "---")
```

```{r}
df_clean <- df %>%
  rename_with( ~ tolower(gsub(" ", "", .x, fixed = TRUE)))%>%
  rename(county=californiacounty) %>% 
  select(censustract, totalpopulation, county, ces4.0percentile, cardiovasculardisease, asthma, pollutionburdenpctl)
```

```{r}
# Aggregate to county level
df_county <- df_clean %>%
  drop_na(ces4.0percentile, cardiovasculardisease) %>%
  filter(totalpopulation > 0) %>%
  group_by(county) %>%
  summarise(
    avg_ces = mean(ces4.0percentile, na.rm = TRUE),
    avg_cvd = mean(cardiovasculardisease, na.rm = TRUE),
    avg_asthma = mean(asthma, na.rm = TRUE),
    total_pop = sum(totalpopulation, na.rm = TRUE)
  )

# Check the top and bottom counties
df_county %>% arrange(desc(avg_ces)) %>% head(5)
df_county %>% arrange(avg_ces) %>% head(5)
```

```{r}
#2.86 to 40.85
state_avg_cvd <- mean(df_clean$cardiovasculardisease, na.rm = TRUE)
state_avg_cvd

p <- ggplot(df_county, aes(x = avg_ces, 
                    y = avg_cvd,
                    size = total_pop)) +
  geom_point(aes(size = total_pop), 
             color = "#1E319E", 
             alpha = 0.6) +
  annotate("text",
         x = 68,
         y = 12.5,
         label = "Los Angeles\n(pop. 10M)",
         size = 3.5,
         hjust = 1,
         color = "#1d3557")+
  scale_size_continuous(range = c(1, 12),
                        labels = scales::comma,
                        name = "Population")+
  geom_hline(yintercept = 13.41,
           linetype = "dashed",
           color = "grey50",
           linewidth = 0.5) +
  annotate("text",
         x = 90,
         y = 13,
         label = "CA average: 13.4",
         size = 3,
         color = "grey50",
         hjust = 1)+
  labs(
    title = "How Does Environmental Burden Relate to Heart Attack Rates?",
    subtitle = "",
    x = "Average CalEnviroScreen 4.0 Percentile (Larger Number, Greater Burden)",
    y = "Average Heart Attack ED Visits\n(age-adjusted rate, per 10,000)",
    caption = "Data: CalEnviroScreen 4.0 (2021) | Population: 2019 ACS estimates
#30DayChartChallenge Day 13 | Created by: Lyndsay Miles"
  ) +
  theme_minimal()+
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    legend.position = "none",
    plot.title = element_text(size = 22),
    plot.subtitle = element_text(size = 14),
    plot.caption = element_text(hjust = 1, size = 12)
  )

p
ggsave("img/cc_2026_day13.png", p,
       width = 10, height = 6, 
       dpi = 300)
```
#Day13 of the #30DayChartChallenge | Topic: Ecosystems 🌎

CalEnviroScreen (CES) ranks California census tracts by pollution burden and population vulnerability. Higher CES percentiles indicate higher impact. This chart explores how the average CES percentile relates to average heart attack ED visit rates by county.

This is a challenge I encounter a lot in public health. We have a lot of information, frequently involving multiple variables and composite scores, and we need to communicate it clearly to a wide audience. I’m not entirely satisfied with this visual; it carries more words and complexity than I’d like.

#Rstats #30DayChartChallenge #DataVisualization #PublicHealth
Data: CalEnviroScreen 4.0 (2021) | Population: 2019 ACS estimates