---
title: "cc_2026_day7"
author: "Lyndsay Miles"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
30 Day Chart Challenge 2026 Day 7: Multiscale

```{r}
library(tidyverse)
library(ggplot2)
library(RColorBrewer)
library(extrafont)
library(showtext)
library(countrycode)

```

```{r}
#PRESS FREEDOM INDEX RANK. Data retrieved from  https://data360.worldbank.org/en/prosperity
df_pfi <- read.csv("data/RWB_PFI_RANK.csv")
```

```{r}
df_pfi_clean <- df_pfi %>% select(REF_AREA, REF_AREA_LABEL, TIME_PERIOD, OBS_VALUE) %>%
  
  rename(
    iso3_code = REF_AREA,
    country   = REF_AREA_LABEL,
    year      = TIME_PERIOD,
    rank      = OBS_VALUE
  )
```

```{r}
df_pfi_clean %>%
  group_by(country) %>%
  summarize(
    rank_2002 = rank[year == min(year)],
    rank_2025 = rank[year == max(year)],
    change    = rank_2025 - rank_2002,  # positive = declined (higher rank number = worse)
    pct_change = ((rank_2025 - rank_2002) / rank_2002) * 100
  ) %>%
  arrange(desc(change))

df_pfi_region <- df_pfi_clean %>%
  filter(!iso3_code %in% c("CSS", "SCG", "XKX", "YUG")) %>%
  mutate(
    region = countrycode(iso3_code,
                         origin      = "iso3c",
                         destination = "region"),
    region = case_when(
      iso3_code == "MEX" ~ "North America",
      TRUE               ~ region
    ),
    is_usa = iso3_code == "USA"
  ) %>%
  mutate(region = gsub("&", "and", region))
```

```{r}
df_pfi_region %>%
  distinct(country, region) %>%
  count(region) 

```

```{r}
#regional average rankings
regional_avg <- df_pfi_region %>%
  group_by(region, year) %>%
  summarize(avg_rank = mean(rank, na.rm = TRUE), .groups = "drop")%>%
  mutate(region = gsub("&", "and", region))

```


```{r}
font_add_google("Playfair Display", "playfair")
showtext_auto()
```

```{r}
ggplot() +
  # Individual country lines  faint
  geom_line(
    data  = df_pfi_region %>% filter(!is_usa),
    aes(x = year, y = rank, group = country),
    color = "#c0c0c0", linewidth = 0.4, alpha = 0.5
  ) +
  # Regional average line
  geom_line(
    data     = regional_avg,
    aes(x = year, y = avg_rank),
    color    = "#3C3B6E", linewidth = 0.75, linetype = "dashed"
  ) +
  # USA line
  geom_line(
    data  = df_pfi_region %>% filter(is_usa),
    aes(x = year, y = rank, group = country),
    color = "#B22234", linewidth = 0.75
  ) +
  scale_y_reverse() +
  scale_x_continuous(breaks = c(2005, 2015, 2025)) +
  facet_wrap(~ region, ncol = 2) +
  labs(
    title    = "Zooming Out: U.S. Press Freedom in a Global Context",
    subtitle = "Grey lines = individual countries | Dashed navy = regional average | Red = United States",
    caption = "Source: Reporters Without Borders Press Freedom Index | #30DayChartChallenge Day 7\nNote: North America regional average includes USA, Canada and Mexico only\nCreated by: Lyndsay Miles",
    x        = NULL,
    y        = "Rank (lower = more press freedom)"
  ) +
  theme_minimal() +
  theme(
    text                  = element_text(family = "playfair"),
    plot.title            = element_text(face = "bold", size = 14, hjust = 0,
                                         margin = margin(b = 6)),
    plot.subtitle         = element_text(size = 9, hjust = 0,
                                         margin = margin(b = 16)),
    plot.caption          = element_text(size = 7, color = "#2d2d2d", hjust = 0,
                                         margin = margin(t = 12)),
    plot.title.position   = "plot",
    plot.caption.position = "plot",
    plot.background       = element_rect(fill = "#ffffff", color = NA),
    panel.background      = element_rect(fill = "#ffffff", color = NA),
    panel.grid.minor      = element_blank(),
    panel.grid.major      = element_line(color = "#f0f0f0"),
    strip.text            = element_text(face = "bold", size = 10),
    axis.text             = element_text(size = 7),
    plot.margin           = margin(20, 20, 20, 20)
  )
```
Day 7 of the #30DayChartChallenge (Multiscale). Staying with the Reporters Without Borders dataset, today I'm zooming out to look at U.S. Press Freedom Rank in a regional and global context.

Each panel shows individual country press freedom trends alongside the regional average (dashed line). The red line is the United States in the North America panel.
