---
title: "cc_2026_day3"
author: "Lyndsay Miles"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
30 Day Chart Challenge 2026 Day 3: Mosaic
Topic: Global Health Spending (tidytuesday): Who pays for healthcare in North America? Mosaic showing financing schemes by country

```{r}
library(tidyverse)
library(ggplot2)
library(RColorBrewer)
library(extrafont)
library(showtext)
```

```{r}
spending_purpose <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2026/2026-04-21/spending_purpose.csv')
financing_schemes <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2026/2026-04-21/financing_schemes.csv')
health_spending <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2026/2026-04-21/health_spending.csv')
```
Cleaning 
```{r}
df <- financing_schemes %>% 
  filter(iso3_code %in% c("USA", "MEX", "CAN"),
         unit == "% of current health expenditure",
         year %in% c(2019:2024),
         value > 0) %>% 
  group_by(iso3_code, financing_scheme) %>% 
  summarize(avg_perc = mean(value), .groups = "drop") %>%
  mutate(
    financing_scheme = case_when(
      grepl("Government", financing_scheme)           ~ "Government/\n compulsory",
      grepl("out-of-pocket", financing_scheme, 
            ignore.case = TRUE)                       ~ "Out-of-pocket",
      grepl("Voluntary", financing_scheme)            ~ "Voluntary insurance",
      TRUE                                            ~ "Other"
    ),
    iso3_code = factor(iso3_code, levels = c("MEX", "CAN", "USA"))
  ) 


mosaic_matrix <- df %>%
  select(iso3_code, financing_scheme, avg_perc) %>%
  tidyr::pivot_wider(names_from = iso3_code, 
                     values_from = avg_perc) %>%
  tibble::column_to_rownames("financing_scheme") %>%
  as.matrix()
```

```{r}
# Plot
library(showtext)
font_add_google("Atkinson Hyperlegible", "atkinson")
showtext_auto()

par(family = "atkinson", cex.main = 1.5, bg = "#f5ebe0")
colnames(mosaic_matrix) <- c("Canada", "Mexico", "United States")
mosaicplot(
  t(mosaic_matrix),
  main     = "How is Health Care Paid for in North America?",
  sub      = "Source: WHO GHED, #tidytuesday | #30DayChartChallenge Day 3\nCreated by: Lyndsay Miles",
  color    = c("#d8b365", "#f5f5f5", "#5ab4ac"),  # swap last two
  border   = "white",
  cex.axis = 0.9,
  las      = 1
)
```
For day 3 of the #30DayChartChallenge, I got into the financing scheme data from this week's #tidytuesday. Mosaic is not something I commonly use, and I wanted to continue to build in R. In the process I learned that the package I've used in the past, ggmosaic(), isn't available in the version of R that I'm running, so I decided to go old school and build this in base R instead.

The chart shows a striking difference in how Mexico finances health care compared to its neighbors, with over 40% of Mexico's health expenditure coming directly from household out-of-pocket payments.