---
title: "cc_2026_day2"
author: "Lyndsay Miles"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
30 Day Chart Challenge 2026 Day 2: Pictogram
Topic: Global Health Spending (tidytuesday)
We're going to build on Day 1's visualization by calculating the per-capita spending on each spending type for USA, UK, Canada, Germany
```{r}
library(tidyverse)
library(ggplot2)
library(RColorBrewer)
library(extrafont)
library(showtext)
```

```{r}
spending_purpose <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2026/2026-04-21/spending_purpose.csv')
```
Cleaning 
```{r}
category_order <- c("Curative care", 
                     "Ancillary services (non-specified by function)",
                     "Long-term care",
                     "Medical goods",
                     "Other health care",
                     "Rehabilitative care",
                     "Governance & admin",
                     "Preventive care")

spending_purpose_na <- spending_purpose %>%
  filter(country_name %in% c("United States of America", "Canada", "Mexico"), 
         unit == "% of current health expenditure", 
         #selected 2022 because this is the most recent year that had data for all 3 countries
         year == "2022") %>%
  group_by(country_name, spending_purpose) %>%
  summarize(pct_raw = sum(value)) %>%
  mutate(
    spending_purpose = recode(spending_purpose,
                              "Governance, and health system and financing administration" = "Governance & admin",
                              "Medical goods (non-specified by function)" = "Medical goods",
                              "Other health care services not elsewhere classified (n.e.c.)" = "Other health care",
                              "Long-term care (health)"   = "Long-term care"
    ),
    country_name = if_else(country_name == "United States of America", "United States", country_name)
  )
```
Add dyslexic-friendly fonts
```{r}
font_add_google("Atkinson Hyperlegible", "atkinson")
showtext_auto()
```
Plot
```{r}
pictogram_data <- spending_purpose_na %>%
  mutate(
    country_name = factor(country_name,
                          levels = c("United States", "Canada", "Mexico"))
  ) %>%
  group_by(country_name) %>%
  mutate(pct_rescaled = pct_raw) %>%
  ungroup()

# --- Define color order left to right ---
category_order <- c("Curative care",
                     "Ancillary services (non-specified by function)",
                     "Long-term care",
                     "Medical goods",
                     "Other health care",
                     "Rehabilitative care",
                     "Governance & admin",
                     "Preventive care")

pictogram_grid <- pictogram_data %>%
  mutate(
    icons_floor = floor(pct_rescaled / 100 * 20),
    remainder   = (pct_rescaled / 100 * 20) - icons_floor
  ) %>%
  group_by(country_name) %>%
  arrange(country_name, desc(remainder)) %>%
  mutate(
    extra      = as.integer(row_number() <= (20 - sum(icons_floor))),
    icon_count = icons_floor + extra
  ) %>%
  ungroup() %>%
  mutate(spending_purpose = factor(spending_purpose, levels = category_order)) %>%
  arrange(country_name, spending_purpose) %>%
  select(country_name, spending_purpose, icon_count)

grid_df <- pictogram_grid %>%
  uncount(icon_count) %>%
  group_by(country_name) %>%
  mutate(
    icon_num = row_number() - 1,
    x        = icon_num %% 20,
    y        = 0
  ) %>%
  ungroup() %>%
  mutate(
    country_name = factor(country_name,
                          levels = c("Mexico", "Canada", "United States")),
    spending_purpose = factor(spending_purpose, levels = category_order)
  )

# --- Plot ---
ggplot(grid_df, aes(x = x, y = country_name, color = spending_purpose)) +
  geom_point(size = 8, shape = 16) +
  scale_color_manual(
    values = c(
      "Curative care"                                  = "#a6611a",
      "Governance & admin"                             = "#80cdc1",
      "Preventive care"                                = "#018571",
      "Ancillary services (non-specified by function)" = "#d0d0d0",
      "Long-term care"                                 = "#d0d0d0",
      "Medical goods"                                  = "#d0d0d0",
      "Other health care"                              = "#d0d0d0",
      "Rehabilitative care"                            = "#d0d0d0"
    ),
    breaks = c("Curative care", "Governance & admin", "Preventive care"),
    labels = c("Curative care", "Governance & admin", "Preventive care")
  ) +
  labs(
    title    = "How U.S. Health Spending Compares to its North American Neighbors",
    subtitle = "Each circle = 5% of total health spending | Grey = all other categories",
    caption  = "Source: WHO Global Health Expenditure Database (GHED) | #30DayChartChallenge Day 2 | #tidytuesday\nCreated by: Lyndsay Miles",
    color    = NULL
  ) +
  theme_void() +
  theme(
    text                  = element_text(family = "atkinson"),
    plot.title            = element_text(face = "bold", size = 14, hjust = 0,
                                         margin = margin(b = 6)),
    plot.subtitle         = element_text(size = 10, hjust = 0,
                                         margin = margin(b = 16)),
    plot.caption          = element_text(size = 7, color = "gray50", hjust = 0,
                                         margin = margin(t = 12)),
    plot.title.position   = "plot",
    plot.caption.position = "plot",
    plot.background       = element_rect(fill = "#f5ebe0", color = NA),
    axis.text.y           = element_text(size = 10, face = "bold", hjust = 1,
                                         margin = margin(r = 8)),
    legend.position       = "bottom",
    legend.text           = element_text(size = 9),
    plot.margin           = margin(20, 20, 20, 20)
  ) +
  guides(color = guide_legend(override.aes = list(size = 5)))
```

For Day 2 of the #30DayChartChallenge (Pictogram) I wanted to see how U.S. % of current health expenditure compares to that of Canada and Mexico using 2022 data from the WHO Global Health Expenditure Database (#tidytuesday).

It is striking how similar all three countries are in terms of Preventive Care spending. In fact, when I switched to 10% per circle, preventive care disappeared entirely for the U.S. and Mexico.

#Rstats #ggplot2